---
title: "EDA"
author: "Marco Tello"
date: "2024-02-22"
output: github_document
---

```{r setup, include=FALSE}
# Differential expression tools
library(edgeR)
# Data format and ID conversion
library(data.table)
library(tidyverse)
# Plot generation 
library(ggplot2)
library(factoextra)
library(pheatmap)

data_path <- "~/GitHub/RNAseq_PECs/data"

knitr::opts_chunk$set(echo = TRUE)
theme_set(theme_bw())
```

# Exploratory Data Analysis

The goal is to explore if:

-   Are there any outlier samples?

-   Are the count distributions are similar across all samples?

-   Do gene expression variability aligns with experimental conditions?

## Count distribution

```{r read data}
# Read count table
# We remove last sample which is not properly identified
count_df <- read.table(file.path(data_path, "Salmon_EstCount_ENSG.tsv"), 
                       header = TRUE) 

count_df %>%
  head()
```

By looking at the density plots of the expression data it is possible to determine if any sample is a clear outlier.

```{r}
count_df %>%
  pivot_longer(!ENSG) %>%
  ggplot(aes(x=value, color = name)) +
    geom_density() +
    scale_x_continuous(trans = 'log10', labels = scales::comma)

```

```{r}
temp <- count_df %>%
  pivot_longer(!ENSG) %>%
  mutate(condition = str_replace(string = name, 
                                 pattern = "_\\d", 
                                 replacement = "")) %>%
  mutate(condition = case_when(
    condition == "Ctrl" ~ "blue",
    condition == "GEN9" ~ "red",
    condition == "H2O2" ~ "darkgreen",
    condition == "PNS2" ~ "purple",
    condition == "PTS3" ~ "pink"))


temp %>%
  ggplot(aes(x=value, color= name))+
  geom_density() +
  scale_x_continuous(trans = 'log10', labels = scales::comma) +
  scale_color_manual(values = temp$condition, 
                     breaks = temp$name) 

```

We can observe that one control sample and one sample from the Gen9 group have slightly more counts than other samples.

### Filter lowly expressed genes

To increase data quality we removed lowly expressed genes in two steps:

1.  Removed genes with no counts across all samples.

```{r Filter no exp genes}
# Remove columns with gene information 
exp_mat <- count_df %>%
  column_to_rownames("ENSG")
# Remove rows with all entries equal zero
exp_mat <- exp_mat[!(rowSums(exp_mat == 0) == ncol(exp_mat)),]

```

2.  Removed genes with less than two reads in more than 12 samples.

```{r filter lowly exp genes}
# We use the average 1st quartile as threshold (2.1), round down to 2
# floor(mean(apply(exp_mat, MARGIN = 2, quantile, 0.25)))

# Select genes to keep
min_reads <- 1
min_samples <- 12
genes_to_keep <- apply(exp_mat >= min_reads, 
                       MARGIN = 1, sum) > min_samples

# Final gene count matrix
exp_mat <- exp_mat[genes_to_keep,]

```

Now we repeat the previous visualizations to explore how the count distributions changed after filtering

```{r warning=FALSE}
exp_mat %>%
  rownames_to_column("ENSG") %>%
  pivot_longer(!ENSG) %>%
  ggplot(aes(x=value, color = name)) +
    geom_density() +
    scale_x_continuous(trans = 'log10', labels = scales::comma)

```

```{r}
temp <- exp_mat %>%
  rownames_to_column("ENSG") %>%
  pivot_longer(!ENSG) %>%
  mutate(condition = str_replace(string = name, 
                                 pattern = "_\\d", 
                                 replacement = "")) %>%
  mutate(condition = case_when(
    condition == "Ctrl" ~ "blue",
    condition == "GEN9" ~ "red",
    condition == "H2O2" ~ "darkgreen",
    condition == "PNS2" ~ "purple",
    condition == "PTS3" ~ "pink"))


temp %>%
  ggplot(aes(x=value, color= name))+
  geom_density() +
  scale_x_continuous(trans = 'log10', labels = scales::comma) +
  scale_color_manual(values = temp$condition, 
                     breaks = temp$name) 
```

This step made samples distributions to look mostly normal on the log10 scale.

## CPM normalization

Before proceeding to further analysis we need to normalize the gene counts, we will use *counts per million* (CPM) to adjust for library size and transform it to log space for better visualization.

```{r cpm norm}
# CPM normalization and transform to log2
expr_log2cpm <- cpm(exp_mat, 
                    log = TRUE, 
                    prior.count = 1) %>% 
  data.frame() 

expr_log2cpm %>%
  rownames_to_column(var = "GeneSymbol") %>%
  pivot_longer(!GeneSymbol) %>%
  dplyr::select(!GeneSymbol) %>%
  ggplot(aes(x=value, color= name))+
  geom_density() 
```

### Violin plots

To directly compare sample gene expression distribution without overlap between density plots we generated the respective violin plots

```{r violin plot}

temp <- expr_log2cpm %>%
  rownames_to_column(var="GeneSymbol") %>%
  pivot_longer(!GeneSymbol) %>%
  dplyr::select(!GeneSymbol) %>%
  mutate(condition = str_replace(string = name, 
                                 pattern = "_\\d+", 
                                 replacement = "")) 
# Convert condition to factor 
temp <- temp %>%
  mutate(condition = factor(condition, 
                            levels =c("Ctrl","GEN9","H2O2","PNS2","PTS3")))

# Plot using ggplot with correct fill mapping
temp %>%
  ggplot(aes(x=name, y=value, fill = condition)) +
  geom_violin()+
  scale_fill_manual(values = c("Ctrl" = "blue", 
                               "GEN9" = "red", 
                               "H2O2" = "darkgreen", 
                               "PNS2" = "purple",
                               "PTS3" = "pink")) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

```

Utilizing log10(CPM) values we can observe that all samples have similar expression distributions.

### Sample-sample correlation plot

Using normalized counts we generated a plot to observe if there is correlation within diet groups which would group the samples accordingly.

```{r sample-sample}
# Annotation

# Standardize conditon names
condition <- names(expr_log2cpm) %>% 
  str_replace(pattern = "_\\d+", 
              replacement = "")

annot <- data.frame(Diet = as.factor(condition),
                    row.names = names(expr_log2cpm))

annot_colors <- list(Diet = c("Ctrl" = "blue", 
                               "GEN9" = "red", 
                               "H2O2" = "darkgreen", 
                               "PNS2" = "purple",
                               "PTS3" = "pink"))

# Heatmap            
expr_log2cpm %>%
  cor() %>%
  pheatmap(annotation_col = annot,
           annotation_row = annot,
           show_rownames = FALSE, 
           annotation_colors = annot_colors, 
           angle_col = 45)


```

### Principal Component Analysis (PCA)

Another way of determining relationship between the samples is through a PCA analysis, which reduces the dimentionality of our data to a set of independent variables (principal components) that represent the major proportion of variability in our data.

```{r PCA CPM}

PCs <- prcomp(t(cpm(exp_mat)), center = TRUE, scale = TRUE)
# Scree plot 
fviz_eig(PCs)
# Scatter plot
eig_val <- get_eigenvalue(PCs)
PCs <- cbind(annot, PCs$x[,1:10])
PCs$sample_id <- rownames(PCs)

PCs <- PCs %>%
  mutate(color_class = case_when(
    condition == "Ctrl" ~ "blue",
    condition == "GEN9" ~ "red",
    condition == "H2O2" ~ "darkgreen",
    condition == "PNS2" ~ "purple",
    condition == "PTS3" ~ "pink")) 

PCs %>%
  ggplot(aes(x = PC1, y = PC2, 
             color = condition)) + 
  geom_point(aes(size = 8)) + 
  scale_color_manual(values = PCs$color_class, 
                     breaks = PCs$condition) +
  labs(x= paste("PC1 (",round(eig_val$variance.percent[1], 2),"%)", sep = ""),
       y= paste("PC2 (",round(eig_val$variance.percent[2], 2),"%)", sep = ""))+
  guides(size = "none") +
  theme_bw()

```

#### Clustering and PCA on top variable genes

For scenarios where experimental variability is expected to be low, it is worth looking at genes with the highest overall variability. If we expect a few differentially expressed genes, the signal is likely to get lost among many genes with no expression changes.

```{r}
temp <- exp_mat

#Top 1000 variable
index <- which( rownames(temp) %in% names(sort(apply(X = temp, MARGIN = 1, var), decreasing = TRUE)[1:1000]))
temp <- temp[index,]

PCs <- prcomp(t(cpm(temp)), center = TRUE, scale = TRUE)
# Scree plot 
fviz_eig(PCs)
# Scatter plot
eig_val <- get_eigenvalue(PCs)
PCs <- cbind(annot, PCs$x[,1:10])
PCs$sample_id <- rownames(PCs)

PCs <- PCs %>%
  mutate(color_class = case_when(
    condition == "Ctrl" ~ "blue",
    condition == "GEN9" ~ "red",
    condition == "H2O2" ~ "darkgreen",
    condition == "PNS2" ~ "purple",
    condition == "PTS3" ~ "pink")) 

PCs %>%
  ggplot(aes(x = PC1, y = PC2, 
             color = Diet)) + 
  geom_point(aes(size = 8)) + 
  scale_color_manual(values = PCs$color_class, 
                     breaks = PCs$Diet) +
  labs(x= paste("PC1 (",round(eig_val$variance.percent[1], 2),"%)", sep = ""),
       y= paste("PC2 (",round(eig_val$variance.percent[2], 2),"%)", sep = ""))+
  guides(size = "none") +
  theme_bw()

```

```{r}
# Heatmap            
temp %>%
  cor() %>%
  pheatmap(annotation_col = annot,
           annotation_row = annot,
           show_rownames = FALSE, 
           annotation_colors = annot_colors, 
           angle_col = 45)

```
